A Dependently Typed Multi-Stage Calculus

### 2019-08-14 text/introduction.tex
Reflected comments on typing from reviwer 1,

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and espace, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
code types per by themselves do not guarantee that the execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem, developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values, and proved a type soundness theorem (for a formal calculus \(\lambda^\alpha\) modeling a pure subset of MetaOCaml).

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.
Section~\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.  For reference, we give the full definition of \LMD and
 proofs of properties in Appendix.

>>>>

MetaOCaml~\cite{calcagno2003implementing,oleg2014} is an extension of
OCaml\footnote{\url{http://ocaml.org}} with special constructs for
multi-stage programming, including brackets and escape, which are
(hygienic) quasi-quotation, and \texttt{run}, which is similar to
\texttt{eval} in Lisp, and cross-stage persistence
(CSP)~\cite{MetaML}.  Programmers can easily write code generators by
using these features.  Moreover, MetaOCaml is equipped with a powerful
type system for safe code generation and execution.  The notion of
code types is introduced to prevent code values that represent
ill-typed expressions from being generated.  For example, a quotation
of expression \texttt{1 + 1} is given type \texttt{int code}
and a code-generating function, which takes a code value \(c\) as an
argument and returns \(c \texttt{ + } c\), is given type \texttt{int
  code -> int code} so that it cannot be applied to, say, a quotation
of \texttt{"Hello"}, which is given type \texttt{string
  code}.  Ensuring safety for \verb|run| is more challenging because
code types per by themselves do not guarantee that the execution of code values never
results in unbound variable errors.  Taha and
Nielsen~\cite{taha2003environment} introduced the notion of
environment classifiers to address the problem, developed a type
system to ensure not only type-safe composition but also type-safe
execution of code values, and proved a type soundness theorem (for a formal calculus \(\lambda^\alpha\) modeling a pure subset of MetaOCaml).

@@

The organization of this paper is as follows.
Section~\ref{sec:informal-overview} gives an informal overview of
\LMD.  Section~\ref{sec:formal-definition} defines \LMD and
Section~\ref{sec:properties} shows properties of \LMD.
Section~\ref{sec:related-work} discusses related work and Section
\ref{sec:conclusion} concludes the paper with discussion of future
work.  For reference, we give the full definition of \LMD and
 proofs of properties in the Appendix.

### 2019-08-14 text/properties.tex
Fixing proofs.
Reflect comments from John.
Reflected comments on typing from reviwer 1,

In this section, we show basic properties of \LMD: preservation, strong normalization, confluence for full reduction, and progress for staged reduction.

@@

Substitution Lemma in \LMD{} is a little more complicated than usual
because there are eight judgment forms and two kinds of substitution.
Term Substitution Lemma states that term substitution $[z \mapsto M]$
preserves derivability of judgments and Stage Substitution Lemma
states similarly for stage substitution $[\alpha\mapsto A]$.

@@

\begin{lemma}[Stage Substitution]
	If $\G \V \mathcal{J}$, then $\G[\beta\mapsto B] \V \mathcal{J}[\beta\mapsto B]$.  Similarly, if $\G \V \mathcal{J}$, then $\V \G[\beta\mapsto B]$.
\end{lemma}

@@

The following Inversion Lemma is needed to prove main theorems.
As usual~\cite{TAPL}, Inversion Lemma enables us to infer the types of subterms of a term from the type of the term.

@@

\begin{proof}
  Each item is strengthen by statements about type equivalence.
  For example, the first statement is augmented by
  \begin{quotation}
    If $\G \V \rho \E (\Pi x:\sigma.\tau) : K @A$, then there exist
    $\sigma'$ and $\tau'$ such that $\rho = \Pi x:\sigma'.\tau'$ and
    $\G \V \sigma \E \sigma' : K @A$ and
    $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
  \end{quotation}
  and its symmetric version.  Then, they are proved simultaneously by induction on derivations.
  Similarly for the others.
 \qed
\end{proof}

>>>>

In this section, we show the basic properties of \LMD: preservation, strong
normalization, confluence for full reduction, and progress for staged
reduction.

@@

The Substitution Lemma in \LMD{} is a little more complicated than usual
because there are eight judgment forms and two kinds of substitution.
The Term Substitution Lemma states that term substitution $[z \mapsto M]$
preserves derivability of judgments. The Stage Substitution Lemma
states similarly for stage substitution $[\alpha\mapsto A]$.

@@

\begin{lemma}[Stage Substitution]
	If $\G \V \mathcal{J}$, then $\G[\beta\mapsto B] \V \mathcal{J}[\beta\mapsto B]$.  Similarly, if $\V \G$, then $\V \G[\beta\mapsto B]$.
\end{lemma}

@@

The following Inversion Lemma is needed to prove the main theorems.
As usual~\cite{TAPL}, Inversion Lemma enables us to infer the types of subterms of a term from the type of the term.

@@

\begin{proof}
  Each item is strengthened by statements about type equivalence.
  For example, the first statement is augmented by
  \begin{quotation}
    If $\G \V \rho \E (\Pi x:\sigma.\tau) : K @A$, then there exist
    $\sigma'$ and $\tau'$ such that $\rho = \Pi x:\sigma'.\tau'$ and
    $\G \V \sigma \E \sigma' : K @A$ and
    $\G, x:\sigma@A\V \tau \E \tau' : J @A$.
  \end{quotation}
  and its symmetric version.  Then, they are proved simultaneously by induction on derivations.
  Similarly for the others.
 \qed
\end{proof}
